---
title: "Spring 2019 ARCHY 483 Experimental Assemblages Report"
author: "Ashlee Breedlove"
date: "`r Sys.Date()`"
output: 
    bookdown::word_document2:
      fig_caption: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE,
                      warning = FALSE)
```

## Introduction 

<<The purpose of this report is to study the five assemblages that Professor Cuthbert Calculus collected from the island of Pulau-Pulau Bomba in Sondonesia. Each assemblage is an undated surface scatter, and they were probably made during the Holocene by the same group of people. The notes were confiscated by Bordurian agents and consequently unavailable. The island was inhabited by forager societies that were extinct when Professor Calculus collected the assemblages. I am here to study the artifacts in advance of a resort being built on the site.>>

<<What is the variability within the between the assemblages? 

You should take inspiration directly from the papers we have read in the seminars, and discussed in lectures. Cite those papers here. Consider the broad anthropological debates about past human behaviour that we have discussed in class, and link your questions to those. Use the word 'debate' or something very similar. Good questions here would be something like: what is the variability in these assemblages? how can it be explained in terms of mobility and risk management? How does the environment structure the technological system?

Give the plan to get answers to your questions. Outline your report to give the reader a preview of what's coming. Give the reader guidance about interesting high points in your results. State the most important of your results; do not leave your reader in suspense. Key phrases: ‘In this study we sought to investigate...’, ‘Three predictions were tested. First . . .’ and 'In brief, we found that..."

## Methods

<<Measurements from five assemblages were recorded in our spreadsheet. The assemblages varied from seven to 25 pieces each, totalling 74. There were three materials recorded. We labelled each artifact as either debitage, flake, or core. Each of our team members (3 of us) took turns measuring and recording the data. Generally, I (Ashlee) measured artifacts, Meghan inputted, and Alec made the input more correct by creating drop-down menus for variables that fit (everything except mass, width, thickness, dorsal number). We followed best practices as laid out in this article (Broman & Woo, 2018).>>

<<Overall we recorded the following variables: artifact type, mass, platform type, platform thickness, platform width, dorsal scar number, dorsal scar pattern, material, and the 16 numbers needed to calculate index of invasiveness. These were chosen based on an analysis of what measurements are worth taking (Scerri, Gravina, Blinkhorn, & Delagnes, 2015). According to the authors, weight, number of scars, platform width and thickness, platform type, and dorsal scar pattern all give meaningful results. In addition, we chose Index of Invasiveness to measure retouch because we didn't want to measure teeny tiny little edges (Clarkson, 2002). The measurements were decided to take were also quick, which is good because we had limited time.>>

<<These help us answer the questions of INSERT HERE because THESE REASONS WHY.

Justify your choice of variables in your data collection: What questions do they help answer? You need to clearly link your variables to your research questions. Precisely describe your analytical methods in the style of an article in the Journal of Archaeological Science. Be as brief but precise in your descriptions. Do not cite or reproduce the labs that we used for the lab work in class. Do not go overboard with detail and generic text-book puff; follow the style of JAS.

<<3/4 minimum citations referenced.>>

## Results

<!-- look at the rubric to know what to put in here -->

Open with 1 or 2 sentence(s) describing just the one or two most striking findings of your work. Use the words 'striking findings' or similar

The text must emphasize the most important observations and present them in order of decreasing interest, beginning with the main finding, for each method. You must describe obvious similarities and contrasts in the data. If there are none, say so.

Charts are used to summarise and visualize contrasts and similarities in your data and they are immediate (quick to interpret), intraocular (hit the reader between the eyes) and inescapable (the trend is obvious). Describe in your text the similarities and differences, the highest and lowest, the most diverse and least diverse, that kind of thing

Your text must not state the result of every variable that you measured because that is very boring for the reader. Show that you have made a thoughtful choice of the variables to investigate and discuss

Note that I have set every chunk below here to `eval = FALSE`. This means that you can knit this Rmd but no code will be run. When you want to knit and have some code run, delete `eval = FALSE` from the chunk that you want to run. The purpose of this is to allow you to test your work as you go by knitting. If the knitting succeeds, then all your code so far is working!

```{r read-in-the-data, eval = FALSE}
library(tidyverse)
library(cowplot)
library(ggbeeswarm)

# read the data --------------------
data <- readxl::read_excel("raw_data.xlsx")
```

#DON'T FORGET TO DELETE ME

```{r delete-this-chunk, eval = FALSE}
# do this interactively to check how things look, and delete 
# this chunk of code from the Rmd after you have worked 
# out how to clean the data. Put your data cleaning code 
# in the next chunk

# check the data class for each variable
glimpse(data) # check that the variables are dbl and chr as you expect

# check for spelling mistakes in your data, make a note so you can correct
# them in the next chunk. R
data %>% # shortcut to type %>% is cmd/ctr + shift + m
  group_by(artifact_type) %>% # replace raw_material with other chr variables
  tally                      # and re-run these three lines of code to check 
                             # for typos in your chr variables

```

#TIDY DATA

```{r tidy-the-data, eval = FALSE}

data <- 
data %>%
 mutate(dorsal_scar_pattern = ifelse(dorsal_scar_pattern == "subcentripital", 
                                  "subcentripetal", 
                                  dorsal_scar_pattern)) %>%
  replace_with_na(data$platform_width, NA)

# NEED TO CHANGE CHARACTER TO NUMBER --
# platform_width, platform_thickness, ioi's, dorsal_scar_no

ioi_ratios_per_zone_long_df <-
data %>%
 select(matches('i\\d')) %>%
 mutate_at(vars(contains("ioi")), ~parse_number(.)) %>%
 mutate(ioi_mean = rowMeans(.)) %>%
 mutate(assemblage_no = data$assemblage_no)
  
```

### Flakes

```{r tally-the-data-1, eval = FALSE}
# Prepare for plotting some categorical variables
# by tallying them up and storing the tally in
# a new data frame

# plot artefact types per assemblage
# tally up the artefact types per assemblage
data_artifact_type_tally <- 
  data %>% 
  group_by(assemblage_no, artifact_type) %>%  
  tally  %>% 
  filter(!is.na(artifact_type),
         !is.na(assemblage_no))

# tally up the raw material types per assemblage
data_raw_material_type_tally <- 
  data %>% 
  group_by(assemblage_no, raw_material) %>% 
  tally  %>% 
  filter(!is.na(raw_material),
         !is.na(assemblage_no))
```


```{r scatterplot-1, eval = FALSE}
# you do not need to change anything major here
# plot the artefact types per assemblage
  ggplot(artefact_type_tally,
         aes(assemblage,               # you may need to change this variable name to match your data
             n, 
             fill = artefact_type)) +  # and you may need to change this name also
  geom_col() +
  labs(x = "Assemblage",
       y = "Number of artefacts",
       fill = "Artefact type") +
  theme_minimal(base_size = 12) +
  scale_fill_viridis_d() 
```

```{r flake-data-prep, eval = FALSE}
# subset complete flakes only for next metric and technological analysis
cf <- artefacts %>% 
  filter(artefact_type == "cf") %>%           # you may need to change artefact_type
  mutate(assemblage = as.factor(assemblage))  # you may need to change assemblage
```

```{r flake-metrics-scatterplot, eval = FALSE}
# You should explore different versions of this plot
# you are welcome to make multiple, different versions of 
# this plot to include in your report

# scatterplot of three variables, this is a handy way to show the relationship 
# between a small number of variables
ggplot(cf,
       aes(Length_mm,            # explore different combinations of variables 
           Width_mm,             # and keep the most interesting results
           size = Thickness_mm,  # to include in your report
           colour = assemblage))  +
  geom_point() +
  labs(x = "Length (mm)",
       y = "Width (mm)",
       size = "Thickness (mm)") +
  scale_color_viridis_d() +
  theme_minimal(base_size = 12) 
```


```{r flake-metrics-facet-plot, eval = FALSE}
# You don't need to change anything major in this chunk unless we 
# tell you to update it to make it work

# create data frame of all numeric variables except retouch
# we need this to make a facetted plot to efficiently show 
# many variables in one plot
cf_numeric_long <- 
cf %>% 
  select(assemblage,   # you may need to change assemblage
         which(sapply(., class) == "numeric"),    # these three lines will
            -c(paste0("t", 1:8),                  # remove the columns with GIUR
               paste0("T", 1:8))) %>%             # i.e. t1, T1, t2, T2, etc
  gather(variable,                                # if your have cols like i1, i2, i3
         value,                                   # then replace t & T here with i
         -assemblage)  %>% 
  filter(!is.na(value)) %>% 
  filter(!is.na(assemblage)) %>% 
  mutate(assemblage = as.factor(assemblage)) 

# facet plot of boxplots of metric variables 
  ggplot(cf_numeric_long,
         aes(assemblage,
             value)) +
  geom_boxplot() +
  geom_quasirandom(alpha = 0.4) +
  facet_wrap(~ variable,
             scales = "free") +
  labs(x = "Assemblage",
       y = "") +
  theme_minimal(base_size = 12) 
```


```{r flake-technology-facet-plot, eval = FALSE}
# You don't need to change anything major in 
# this chunk unless we tell you to update it 
# to make it work

# plot all categorical variables on complete flakes
  
# make a list of tallies of categorical variables
  cf_list_categorical_variables_tally <- 
    cf %>% 
  select(which(sapply(., class) == "character"), 
         assemblage,        # watch out here... 
         -artefact_type,    # you may need to change these names
         -artefact_number,  # if your column names differ from
         -artefact_id,      # ours
         -raw_material) %>% 
  gather(variable, 
         value, 
         -assemblage) %>% 
  mutate(assemblage = as.factor(assemblage)) %>% 
  group_by(assemblage, 
           variable, 
           value) %>% 
  tally %>% 
  filter(!is.na(value),
         !is.na(assemblage)) %>% 
  ungroup %>% 
  group_split(variable)  

# plot each variable across all assemblages  
  map(cf_list_categorical_variables_tally,
      ~ggplot(.x, aes(assemblage, 
             n,
             fill = value)) +
  geom_col() +
  labs(x = "Assemblage",
       y = "Number",
       fill = .x$variable) +
  scale_fill_viridis_d() +
  theme_minimal(base_size = 12))  %>% 
  plot_grid(plotlist  = .) 
```

### Cores

<!-- copy-paste from flakes, and update, send me a screenshot if you get errors and we'll help you out -->

```{r core-data-prep, eval = FALSE}
# subset complete flakes only for next metric and technological analysis
cores <- artefacts %>% 
  filter(artefact_type == "core") %>%           # you may need to change artefact_type
  mutate(assemblage = as.factor(assemblage))    # you may need to change assemblage
```

### Retouched pieces

```{r retouch-measured-using-the-GIUR, eval = FALSE}
# You don't need to change this chunk unless we tell you to update it to make it work.
# If you recorded columns of t1, T1, t2, T2 etc., then use this chunk, if not,
# delete it and look for the next one
  
# we assume that the data include a set of columns like this
# t1 T1 t2 T2 ... and we select only those columns, then
# we divide pairs of columns to get the ratio for that zone
  
divide_two_cols <- function(df){
 col<-seq(1,ncol(df),by=2)
 df[,col] / df[,-col]
}

giur_ratios_per_zone_long_df <- 
artefacts %>% 
  select(matches('t\\d|T\\d'), 
         assemblage) %>%
  group_split(assemblage, keep = FALSE) %>% 
  map(~divide_two_cols(.x)) %>% 
  set_names(unique(flakes$assemblage)) %>% 
  map_dfr(rowMeans) %>% 
  gather(variable, value) 

  ggplot(giur_ratios_per_zone_long_df, 
         aes(variable, 
             value)) +
  geom_boxplot()  +
  geom_quasirandom(alpha = 0.4) +
  labs(x = "Assemblage",
       y = "GIUR") +
  theme_minimal(base_size = 12) +
  scale_y_continuous(trans='log2')
```


```{r retouch-measured-using-the-IOI, eval = FALSE}
# You don't need to change this chunk unless we tell you to update it to make it work.
# If you recorded columns of i1, i2, i3, etc. then use this chunk
# If you recorded the GIUR, then use the chunk above. If you did neither, 
# let me know so I can help you

ioi_ratios_per_zone_long_df <- 
artefacts %>% 
  select(matches('i\\d')) %>%
  mutate(ioi_mean = rowMeans(.)) %>% 
  mutate(assemblage = artefacts$assemblage) 

  ggplot(ioi_ratios_per_zone_long_df, 
         aes(assemblage, 
             ioi_mean)) +
  geom_boxplot()  +
  geom_quasirandom(alpha = 0.4) +
  labs(x = "Assemblage",
       y = "IOI") +
  theme_minimal(base_size = 12) +
  scale_y_continuous(trans='log2')
```


## Discussion

The first sentence must be an explanation of how findings link to the context of the anthropological problem stated in introduction

Offer some brief explanations of your results and ward off counter-claims. Connect your data to conceptual frameworks from readings we've discussed in class, and cite those readings here.

Describe suprising and unexpected results and offer an explanation for them. List the most substantial limitations to the study that you encountered. Do not go overboard with every little hiccup you encountered.

You need to have a minimum of five citations of relevant work in the Discussion section, they must be very clearly relevant to, and tied deeply into, your discussion. You must draw on papers we've read and discussed in our seminars. They should be papers that you want to use as an analogy, so you want to refer to the details of their interpretation and their data to make an analogy to your own data

## Conclusion

Remind the reader of your aim and the exact questions that you stated in your introduction.

State whether your predictions was supported or contradicted, and connect them to the broader anthropological issues that you mentioned in your introduction, including citations

Recommend specific further research to address unresolved and new questions that arose during your work.

## Collaboration

<<I worked on this with Meghan and Alec.>>

## References cited

Broman, K. W., & Woo, K. H. (2018). Data Organization in Spreadsheets. The American Statistician, 72(1), 2-10.

Clarkson, C. (2002). An Index of Invasiveness for the Measurement of Unifacial and
Bifacial Retouch: A Theoretical, Experimental and Archaeological Verification. Journal of Archaeological Science, 29, 65–75.

Scerri, E. M. L., Gravina B., Blinkhorn, J., & Delagnes, A. (2015). Can Lithic Attribute Analyses Identify Discrete Reduction Trajectories? A Quantitative Study Using Refitted Lithic Sets. Journal of Archaeological Method and Theory, 23(2), 669-691. 

<!-- don't delete the line below, it will count the words in your report -->
Word count: `r wordcountaddin::word_count()`