---
title: "Spring 2019 ARCHY 483 Experimental Assemblages Report"
author: "Ashlee Breedlove"
date: "`r Sys.Date()`"
output: 
    bookdown::word_document2:
      fig_caption: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,     # this ensures we can see your code in the output
                      message = FALSE, # this stops messages from appearing
                      warning = FALSE) # this stops warnings from appearing
```

## Introduction 

<!-- look at the rubric to know what to put in here -->

## Methods

<!-- look at the rubric to know what to put in here -->

## Results

<!-- look at the rubric to know what to put in here -->

Note that I have set every chunk below here to `eval = FALSE`. This means that you can knit this Rmd but no code will be run. When you want to knit and have some code run, delete `eval = FALSE` from the chunk that you want to run. The purpose of this is to allow you to test your work as you go by knitting. If the knitting succeeds, then all your code so far is working!

```{r read-in-the-data, eval = FALSE}
library(tidyverse)
library(cowplot)
library(ggbeeswarm)

# read the data --------------------
artefacts <- readxl::read_excel("Experimental Assemblage Data.xlsx")
```

```{r delete-this-chunk, eval = FALSE}
# do this interactively to check how things look, and delete 
# this chunk of code from the Rmd after you have worked 
# out how to clean the data. Put your data cleaning code 
# in the next chunk

# check the data class for each variable
glimpse(artefacts) # check that the variables are dbl and chr as you expect

# check for spelling mistakes in your data, make a note so you can correct
# them in the next chunk. R
artefacts %>% # shortcut to type %>% is cmd/ctr + shift + m
  group_by(raw_material) %>% # replace raw_material with other chr variables
  tally                      # and re-run these three lines of code to check 
                             # for typos in your chr variables

```


```{r tidy-the-data, eval = FALSE}
# Do not delete this chunk, we want to see how you have cleaned your 
# data to prepare it for analysis
# you may need to clean up some cols, e.g. cols that are chr but should be num
# you may need to fix some typos due to accidental mistakes during data entry
artefacts <- 
artefacts %>%
  mutate(t5 = parse_number(t5)) %>%    # example of how to convert chr to dbl
  separate(artefact_number,            # if you recorded each artefact as 1-1, 1-2
           into = c("assemblage",      # etc., then you need these lines with
                    "artefact_id"),    # separate() that into two columns
           sep = "-",
           remove = FALSE) %>% 
  # these next lines with mutate() show how to correct spelling mistakes in
  # your data
 mutate(Termination_Type = ifelse(Termination_Type == "hinginging", 
                                  "hinging", 
                                  Termination_Type)) %>% 
 mutate(Termination_Type = ifelse(Termination_Type == "feathery", 
                                  "feathered", 
                                  Termination_Type)) 
  
```

### Flakes

```{r tally-the-data-1, eval = FALSE}
# Prepare for plotting some categorical variables
# by tallying them up and storing the tally in
# a new data frame

# tally up the artefact types per assemblage
artefact_type_tally <- 
  artefacts %>% 
  group_by(assemblage,          # The column names in your data may be slightly
           artefact_type) %>%   # different to what I've got here. You need to edit
  tally  %>%                    # the code throughout here to use your column names
  filter(!is.na(artefact_type), # These two lines remove NA values in these two
         !is.na(assemblage))    # columns. If there are no NAs, they do nothing

# tally up the raw material types per assemblage
raw_material_type_tally <- 
  artefacts %>% 
  group_by(assemblage, 
           raw_material) %>% 
  tally  %>% 
  filter(!is.na(raw_material),
         !is.na(assemblage))
```


```{r scatterplot-1, eval = FALSE}
# you do not need to change anything major here
# plot the artefact types per assemblage
  ggplot(artefact_type_tally,
         aes(assemblage,               # you may need to change this variable name to match your data
             n, 
             fill = artefact_type)) +  # and you may need to change this name also
  geom_col() +
  labs(x = "Assemblage",
       y = "Number of artefacts",
       fill = "Artefact type") +
  theme_minimal(base_size = 12) +
  scale_fill_viridis_d() 
```

```{r flake-data-prep, eval = FALSE}
# subset complete flakes only for next metric and technological analysis
cf <- artefacts %>% 
  filter(artefact_type == "cf") %>%           # you may need to change artefact_type
  mutate(assemblage = as.factor(assemblage))  # you may need to change assemblage
```

```{r flake-metrics-scatterplot, eval = FALSE}
# You should explore different versions of this plot
# you are welcome to make multiple, different versions of 
# this plot to include in your report

# scatterplot of three variables, this is a handy way to show the relationship 
# between a small number of variables
ggplot(cf,
       aes(Length_mm,            # explore different combinations of variables 
           Width_mm,             # and keep the most interesting results
           size = Thickness_mm,  # to include in your report
           colour = assemblage))  +
  geom_point() +
  labs(x = "Length (mm)",
       y = "Width (mm)",
       size = "Thickness (mm)") +
  scale_color_viridis_d() +
  theme_minimal(base_size = 12) 
```


```{r flake-metrics-facet-plot, eval = FALSE}
# You don't need to change anything major in this chunk unless we 
# tell you to update it to make it work

# create data frame of all numeric variables except retouch
# we need this to make a facetted plot to efficiently show 
# many variables in one plot
cf_numeric_long <- 
cf %>% 
  select(assemblage,   # you may need to change assemblage
         which(sapply(., class) == "numeric"),    # these three lines will
            -c(paste0("t", 1:8),                  # remove the columns with GIUR
               paste0("T", 1:8))) %>%             # i.e. t1, T1, t2, T2, etc
  gather(variable,                                # if your have cols like i1, i2, i3
         value,                                   # then replace t & T here with i
         -assemblage)  %>% 
  filter(!is.na(value)) %>% 
  filter(!is.na(assemblage)) %>% 
  mutate(assemblage = as.factor(assemblage)) 

# facet plot of boxplots of metric variables 
  ggplot(cf_numeric_long,
         aes(assemblage,
             value)) +
  geom_boxplot() +
  geom_quasirandom(alpha = 0.4) +
  facet_wrap(~ variable,
             scales = "free") +
  labs(x = "Assemblage",
       y = "") +
  theme_minimal(base_size = 12) 
```


```{r flake-technology-facet-plot, eval = FALSE}
# You don't need to change anything major in 
# this chunk unless we tell you to update it 
# to make it work

# plot all categorical variables on complete flakes
  
# make a list of tallies of categorical variables
  cf_list_categorical_variables_tally <- 
    cf %>% 
  select(which(sapply(., class) == "character"), 
         assemblage,        # watch out here... 
         -artefact_type,    # you may need to change these names
         -artefact_number,  # if your column names differ from
         -artefact_id,      # ours
         -raw_material) %>% 
  gather(variable, 
         value, 
         -assemblage) %>% 
  mutate(assemblage = as.factor(assemblage)) %>% 
  group_by(assemblage, 
           variable, 
           value) %>% 
  tally %>% 
  filter(!is.na(value),
         !is.na(assemblage)) %>% 
  ungroup %>% 
  group_split(variable)  

# plot each variable across all assemblages  
  map(cf_list_categorical_variables_tally,
      ~ggplot(.x, aes(assemblage, 
             n,
             fill = value)) +
  geom_col() +
  labs(x = "Assemblage",
       y = "Number",
       fill = .x$variable) +
  scale_fill_viridis_d() +
  theme_minimal(base_size = 12))  %>% 
  plot_grid(plotlist  = .) 
```

### Cores

<!-- copy-paste from flakes, and update, send me a screenshot if you get errors and we'll help you out -->

```{r core-data-prep, eval = FALSE}
# subset complete flakes only for next metric and technological analysis
cores <- artefacts %>% 
  filter(artefact_type == "core") %>%           # you may need to change artefact_type
  mutate(assemblage = as.factor(assemblage))    # you may need to change assemblage
```

### Retouched pieces

```{r retouch-measured-using-the-GIUR, eval = FALSE}
# You don't need to change this chunk unless we tell you to update it to make it work.
# If you recorded columns of t1, T1, t2, T2 etc., then use this chunk, if not,
# delete it and look for the next one
  
# we assume that the data include a set of columns like this
# t1 T1 t2 T2 ... and we select only those columns, then
# we divide pairs of columns to get the ratio for that zone
  
divide_two_cols <- function(df){
 col<-seq(1,ncol(df),by=2)
 df[,col] / df[,-col]
}

giur_ratios_per_zone_long_df <- 
artefacts %>% 
  select(matches('t\\d|T\\d'), 
         assemblage) %>%
  group_split(assemblage, keep = FALSE) %>% 
  map(~divide_two_cols(.x)) %>% 
  set_names(unique(flakes$assemblage)) %>% 
  map_dfr(rowMeans) %>% 
  gather(variable, value) 

  ggplot(giur_ratios_per_zone_long_df, 
         aes(variable, 
             value)) +
  geom_boxplot()  +
  geom_quasirandom(alpha = 0.4) +
  labs(x = "Assemblage",
       y = "GIUR") +
  theme_minimal(base_size = 12) +
  scale_y_continuous(trans='log2')
```


```{r retouch-measured-using-the-IOI, eval = FALSE}
# You don't need to change this chunk unless we tell you to update it to make it work.
# If you recorded columns of i1, i2, i3, etc. then use this chunk
# If you recorded the GIUR, then use the chunk above. If you did neither, 
# let me know so I can help you

ioi_ratios_per_zone_long_df <- 
artefacts %>% 
  select(matches('i\\d')) %>%
  mutate(ioi_mean = rowMeans(.)) %>% 
  mutate(assemblage = artefacts$assemblage) 

  ggplot(ioi_ratios_per_zone_long_df, 
         aes(assemblage, 
             ioi_mean)) +
  geom_boxplot()  +
  geom_quasirandom(alpha = 0.4) +
  labs(x = "Assemblage",
       y = "IOI") +
  theme_minimal(base_size = 12) +
  scale_y_continuous(trans='log2')
```


## Discussion

## Conclusion

## Collaboration

## References cited


<!-- don't delete the line below, it will count the words in your report -->
Word count: `r wordcountaddin::word_count()`